Impact of Demographics and Skills of the Observer on Visual Statistical Inference

### 2013-05-22 sociology_paper.tex
abstract updated

\begin {abstract}  
Visual Inference is dependent on the careful evaluation of the lineups by individual observers. Each individual is different in their cognitive phycology and judiciousness which can affect the power of visual inference. To estimate the power of visual inference this can be controlled by getting evaluations from multiple people from a diverse population. But the other factor that may affect the power of visual inference is the demographics and skills of the observer. In this paper we examine this in details. The simulation experiments suggest that individual skill as well as demographics are very significant for the power of visual inference. Moreover, the data shows the evidence of learning or getting skilled while sequentially evaluating multiple lineups.

@@

Because of its convenience it is getting popular for scientific research study. In comparison with a lab study \cite{suri:2010} perform the same study using MTurk and demonstrate that their study results are as good as the lab study results even though MTurk study required less time, cost and provided more convenience. \cite{majumder:2012} recruited people from MTurk for their simulation study in estimating the power of visual statistical inference. They have done numerous pilot studies in lab before doing actual MTurk study and found similar results. \cite{mason:2012} explains numerous features of MTurk and describes how it can be used as part of human behavioral study.

@@

\subsection{Learning Trend of the Observer} Since each subject is shown multiple lineups, they have a chance for self learning and do better on lineups shown later in the sequence. They may learn from their previous mistake or they may learn about the features of the plots being evaluated as they progress. We examine this by fitting generalized mixed effect model presented in \cite{majumder:2012} with lineups and subjects as random effect and no fixed effect. This should give us the estimation of proportion correct responses adjusted for lineup difficulties and individual skills. The resulting residual should show some trend if plotted against attempt. Figure \ref{fig:learning_trend} shows such plots for three experiments naming experiment 5 through 6. For experiment 5 we see a downward trend in absolute residual as the attempt progress. This may indicate a learning trend but this trend is not statistically significant. We see upward trend for experiment 6 and 7 but again they are not statistically significant too.

>>>>

\begin {abstract}  
Visual Inference is dependent on the careful evaluation of the lineups by individual observers. Each individual is different in their cognitive phycology and judiciousness which can affect the power of visual inference. To estimate the power of visual inference this can be controlled by getting evaluations from multiple people from a diverse population. But the other factor that may affect the power of visual inference is the demographics and skills of the observer. In this paper we examine this in details. The simulation experiments suggest that individual skill as well as demographics are very significant for the power of visual inference. We examine the evidence of learning or getting skilled while sequentially evaluating multiple lineups. The experimental data suggest that there is no learning effect. The location of actual plot in the lineups does not have any significant effect as well.

@@

Because of its convenience it is getting popular for scientific research study. In comparison with a lab study \cite{suri:2010} perform the same study using MTurk and demonstrate that their study results are as good as the lab study results even though MTurk study required less time, cost and provided more convenience. \cite{majumder:2012} recruited people from MTurk for their simulation study in estimating the power of visual statistical inference. They have done numerous pilot studies in lab before doing actual MTurk study and found similar results. \cite{mason:2012} explains various features of MTurk and describes how it can be used as part of human behavioral study.

@@

\subsection{Learning Trend of the Observer} Each subject participating the experiment has seen multiple number of lineups for evaluations. Suppose $K$ be the number of lineups evaluated by a subject and $A_k$ be the $kth$ attempt, $k=1,2, ..., K$. Since each subject is shown multiple lineups, they have a chance for self learning and do better on lineups shown later in the sequence. They may learn from their previous mistake or they may learn about the features of the plots being evaluated as they progress. Thus if there is any learning trend that should be apparent in proportion correct over $A_k$. But to obtain this we have to adjust for lineup difficulty and individual performances.

To accomplish this we fit generalized mixed effect model presented in \cite{majumder:2012} with lineups and subjects as random effect and no fixed effect. This gives the estimation of proportion correct responses  and we obtain residuals of the fitted model. The residuals give variations in proportion correct adjusted for lineup difficulties and individual skills. If there is any learning trend over different attempt $A_k$, the residuals should have that information since $A_k$ was not included as a covariate in the model. At this point we focus on absolute residuals over different attempts to study the deviations due to attempts. An smaller absolute residual at attempt $A_k$ should indicate learning from all other attempts less than $k$.

A plot of absolute residuals against the attempts $A_k$ should reveal the trend if there is any. Figure \ref{fig:learning_trend} shows such plots for three experiments naming experiment 5 through 6. For experiment 5 we see a downward trend in absolute residual as the attempt progress. This may indicate a learning trend but this trend is not statistically significant. We see upward trend for experiment 6 and 7. This could be attributed to the exhaustion of subjects' attempt after some earlier evaluations are made. But again these upward trends are not statistically significant as well.
